Latent Variable Clustering for Classification of Events in the AT-TPC

### 2020-05-22 doc/Manuscript/article.tex
Few edits and comments

\section{Experimental Details} 
The goal of the experiment was to measure the evolution of the elastic and inelastic cross sections between $^{46}$Ar and protons as a function of energy (also called excitation function), and observe resonances in the composite system $^{47}$K that correspond to analog states in the nucleus $^{47}$Ar. Spectroscopic information can then be obtained from the shape and amplitude of the observed resonances \cite{Bradt2018} {\color{blue} check references: first should be Commisioning... this should be physics paper}. 
The experiment was performed at the National Superconducting Cyclotron Facility (NSCL) where a $^{46}$Ar beam was produced via fragmentation of a $^{48}$Ca beam on a $^9$Be target at about 140 MeV/u. The $^{46}$Ar isotopes were then filtered, thermalized, and finally re-accelerated to 4.6 MeV/u by a linear accelerator. This scheme was used to produce a low-emittance beam. which is necessary to guarantee a good energy resolution in the excitation function. Because the  $^{46}$Ar beam particles lose energy as they traverse the target gas volume, the position of the reaction vertex along the beam axis is directly related to the energy at which the reaction occurs. This allows the AT-TPC to measure the excitation function over a wide range of energies from a single beam energy.

@@

\subsection{Data}

@@

The AT-TPC data we studied for this work was originally recorded as charge time-series for each pad on the micromegas.
In this representation an event is a record of 512 time-buckets for each of the $\sim10^4$ detector pads; however, in our analysis, we represent each event as a 2D slice.
We chose to represent the data in 2D to facilitate the use of sophisticated image-recognition machine learning models.
The 2D representation of events was computed by viewing the event as a point cloud of peaks in the charge time-series and summing over the time-axis.
Additionally, we log-transform and normalize the data to  the interval $[0, 1]$ for each point. The final representation for each event is a $128 \times 128$ matrix with each element ranging from $0$ to $1$.

One of the significant considerations for the analysis of AT-TPC data is to inject machine learning methods for track identification at the correct point in the analysis pipeline.
Using raw data is advantageous as it provides an unbiased view of the event, but the data volume and noise levels might be prohibitive for analysis.
On the opposite end of the pipeline, we will incrementally add bias to the analysis the more steps we take before applying the algorithm. With the benefit being that more pre-processing improves the signal-to-noise ratio, possibly improving model performance.
To explore this trade-off between model performance and pre-processing bias, we performed analysis on simulated, cleaned and raw events.

@@

The simulated AT-TPC tracks were simulated with the \lstinline{pytpc} package developed by \citet{Bradt2017a}. Using the same parameters as for the ${}^{46}$Ar$(p, p)$ experiment, a small set of $N=4000$ events were generated per class, as well as a larger set of $N=40000$ events per class.

To emulate the real-data case, we set a subset of the simulated data to be labelled and treat the rest as unlabelled data. We chose this partition to be $15\%$ of each class. We denote this subset and its associated labels as $\gamma_L=(\boldsymbol{X}_L, \boldsymbol{y}_L)$, the entire dataset which we will denote as $\boldsymbol{X}_F$. To clarify, please note that $\boldsymbol{X}_L \subset \boldsymbol{X}_F$.

We display two simulated events in figure \ref{fig:sim_samples}. The top row illustrates a proton-event, and the bottom a carbon-event.

@@

The events analyzed in this section were retrieved from the ${}^{46}$Ar resonant proton scattering experiment recorded with the AT-TPC. 
While we denote these events as raw, it is important to not that what we mean is a raw 2D representation of peak-only events.

@@

We begin by exploring a naive K-means approach on the latent space of a pre-trained network. Subsequently, we investigate other clustering methods and two autoencoder based clustering algorithms, as outlined in section \ref{sec:deep_clustering}.

\todo{Drop this argument or build it in the ML section?}
This chapter builds on the previous results from semi-supervised classification. We observe that we can construct high-quality latent spaces. These high-quality spaces facilitate an investigation of clustering techniques.

\todo{Reword this to not contrast against semi-supervised}
The approach for clustering of events is different from the semi-supervised approach in two meaningful ways. First, it is a harder task, as we will demonstrate. The clustering task thus necessitates a more exploratory approach to the problem. Second, as a consequence of the challenge, the focus will be a bit different than for the semi-supervised approach. We will still utilize the same architectures and models starting with a search over the parameter space over which we measure the performance using the adjusted rand score (ARS) and accuracy defined in section \ref{sec:unsupervised_perf} and \ref{sec:supervised_perf}, respectively.

As with chapter  \ref{chap:classification} where we explored the semi-supervised results, we begin this chapter by considering the VGG16 pre-trained model as a benchmark.

Lastly, we note that the focus of this work is mainly on discovering possible avenues for further research. This focus requires a broad scan of possible avenues rather than a rigorous analysis of one specific model.
\subsection{Transfer learning}
As in chapter \ref{ch:architectures}, we also use the VGG16 pre-trained network as a baseline for the clustering performance. We begin by considering a classical K-means approach to clustering. However, the output from the VGG16 network is very high dimensional with output vectors in $\R^{8192}$. \todo{reword} One of the primary concerns is then the curse of dimensionality, where the ratio of distances goes to one with increasing dimensionality as shown by \citet{Aggarwal}. However, one of the central caveats to the authors  finding is that the elements are uniformly distributed in the latent space. It is then possible that all the class information lies in some sub-space of the latent data. To investigate this, we perform clustering analysis using the full representation and the $10^2$ first principal components only.

@@

We begin by investigating the K-means clustering algorithm on the VGG16 latent space. As in chapter \ref{chap:classification} the VGG16 model is pre-trained on the imagenet dataset creating a set of vectors $\boldsymbol{x} \in \R^{8192}$. To cluster we use \lstinline{scikit-learn} implementation of the K-means algorithm, with default parameters \cite{Pedregosa2011}. The results of the clustering runs are included in table \ref{tab:clstr_vgg}. We observe that we are able to attain near-perfect clustering on simulated data and that there is a sharp decline in performance as we add noise by moving to the filtered and full datasets.

@@

We repeat this analysis using a PCA dimensionality reduction on the latent space of the VGG16 model. This is done to estimate to what degree the class separating information is encoded in the entirety of the latent space, or in some select regions. The results from the PCA analysis were virtually identical to the results sans the PCA, and so we omit them for brevity.

@@

In addition to the results presented in this section, we performed clustering with a number of different algorithms included in the \lstinline{scikit-learn} package. None of them provided any notable differences from the K-means results or were significantly worse. Notably, the DBSCAN algorithm failed to provide any useful clustering results. We find this important as one of the significant drawbacks of K-means, and the deep clustering algorithms presented in section \ref{sec:deep_clustering}, is that they are all dependent on pre-determining the number of clusters. This is not the case for DBSCAN. 
\subsection{Autoencoder clustering}
\todo{look at grayscale images or colorbar with propto charge}

>>>>

\section{Experimental Details} 
The goal of the experiment was to measure the evolution of the elastic and inelastic cross sections between $^{46}$Ar and protons as a function of energy (also called excitation function), and observe resonances in the composite system $^{47}$K that correspond to analog states in the nucleus $^{47}$Ar. Spectroscopic information can then be obtained from the shape and amplitude of the observed resonances \cite{Bradt2018} {\color{blue} check references: first should be Commisioning... this should be physics paper}. 
The experiment was performed at the National Superconducting Cyclotron Facility (NSCL) where a $^{46}$Ar beam was produced via fragmentation of a $^{48}$Ca beam on a $^9$Be target at about 140 MeV/u. The $^{46}$Ar isotopes were then filtered, thermalized, and finally re-accelerated to 4.6 MeV/u by a linear accelerator. This scheme was used to produce a low-emittance beam, which is necessary to guarantee a good energy resolution in the excitation function. Because the  $^{46}$Ar beam particles lose energy as they traverse the target gas volume, the position of the reaction vertex along the beam axis is directly related to the energy at which the reaction occurs. This allows the AT-TPC to measure the excitation function over a wide range of energies from a single beam energy.

@@

\subsection{Data}
{\color{green}What is the intent of this section? Display some data examples to illustrate the points above?}

@@

The AT-TPC data we studied for this work was originally recorded as charge time-series for each pad on the micromegas {\color{green} We haven't defined what micromegas is}.
In this representation an event is a record of 512 time-buckets for each of the $\sim10^4$ detector pads; however, in our analysis, we represent each event as a 2D slice.
We chose to represent the data in 2D to facilitate the use of sophisticated image-recognition machine learning models.
The 2D representation of events was computed by viewing the event as a point cloud in the charge time-series and summing over the time-axis.
Additionally, we log-transform and normalize the data to  the interval $[0, 1]$ for each point. The final representation for each event is a $128 \times 128$ matrix with each element ranging from $0$ to $1$.

One of the significant considerations for the analysis of AT-TPC data is to inject machine learning methods for track identification at the best point in the analysis pipeline.
Using raw data is advantageous as it provides an unbiased view of the event, but the data volume and noise levels might be prohibitive for analysis.
On the opposite end of the pipeline, we will incrementally add bias to the analysis the more steps we take before applying the algorithm. With the benefit being that more pre-processing improves the signal-to-noise ratio, possibly improving model performance.
To explore this trade-off between model performance and pre-processing bias, we performed analysis on simulated, cleaned and raw events.

@@

AT-TPC tracks were simulated with the \lstinline{pytpc} package developed by \citet{Bradt2017a}. Using the same parameters as for the ${}^{46}$Ar$(p, p)$ experiment, a small set of $N=4000$ events per class was generated, as well as a larger set of $N=40000$ events per class.

To emulate the real-data case, we set a subset of the simulated data to be labelled and treat the rest as unlabelled data. We choose this partition to be $15\%$ of each class. We denote this subset and its associated labels as $\gamma_L=(\boldsymbol{X}_L, \boldsymbol{y}_L)$, while the entire dataset which is identified as $\boldsymbol{X}_F$. Note that $\boldsymbol{X}_L \subset \boldsymbol{X}_F$.

Two simulated events are displayed in figure \ref{fig:sim_samples}. The top row illustrates a proton-event, and the bottom a carbon-event.

@@

The events analyzed in this section were retrieved from the ${}^{46}$Ar resonant proton scattering experiment recorded with the AT-TPC. 
While we denote these events as raw, it is important to note that what we mean is a raw 2D representation of peak-only events.

@@

We begin by exploring a naive K-means approach on the latent space of a pre-trained network. Subsequently, we investigate other clustering methods and the autoencoder based clustering algorithm outlined in section \ref{sec:mixae}.

\subsection{Transfer learning}
Building on the work by \citet{Kuchera2019}, we also use the VGG16 pre-trained network as a baseline for the clustering performance. We begin by considering a classical K-means approach to clustering. However, the output from the VGG16 network is very high dimensional which traditionally conficts with good K-means results. With output vectors in $\R^{8192}$ the ratio between $L_p$ distances goes to one with increasing dimensionality, as shown by \citet{Aggarwal}. However, one of the central assumptions in the authors' finding is that the elements are uniformly distributed in the latent space. They note that it is then possible that if the class information lies in some sub-space of the latent data, $L_p$-norms may provide useful clusterings even in high dimensional spaces. To investigate this, we perform a $L_2$-norm based K-means clustering using the full VGG16 representation and the $10^2$ first principal components only.

@@

We begin by investigating the K-means clustering algorithm on the VGG16 latent space. The VGG16 model is pre-trained on the imagenet dataset, and subsequently applied to the AT-TPC events without the classification layers. This creates a set of vectors $\boldsymbol{x} \in \R^{8192}$ as our new representation of the events. To cluster we use \lstinline{scikit-learn} implementation of the K-means algorithm, with default parameters \cite{Pedregosa2011}. The results of the clustering runs are included in table \ref{tab:clstr_vgg}. We observe that we are able to attain near-perfect clustering on simulated data and that there is a sharp decline in performance as we add noise by moving to the filtered and raw datasets.

@@

We repeat this analysis using a PCA dimensionality reduction {\color{green}(please define)} on the latent space of the VGG16 model. This is done to estimate to what degree the class separating information is encoded in the entirety of the latent space, or in some select regions. The results from the PCA analysis were virtually identical to the results sans the PCA, and so we omit them for brevity.

@@

In addition to the results presented in this section, we performed clustering with a number of different algorithms included in the \lstinline{scikit-learn} package. None of them provided any notable differences from the K-means results or were significantly worse. Notably, the DBSCAN algorithm {\color{green} (add reference)} failed to provide any useful clustering results. We find this important as one of the significant drawbacks of K-means, and the deep clustering algorithms presented in section \ref{sec:deep_clustering}, is that they are all dependent on pre-determining the number of clusters. This is not the case for DBSCAN. 
\subsection{Autoencoder clustering}
\todo{look at grayscale images or colorbar with propto charge}
